---
title: "Quality Control with SingleCellTools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quality Control with SingleCellTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**Package:** `SingleCellTools`  
**Author:** K. Garrett Evensen, PhD — Bioinformatics Analyst II, The Salk Institute  
**Version:** 2.8  
**Source:** [github.com/gevensen95/SingleCellTools](https://github.com/gevensen95/SingleCellTools)

---

## Table of Contents

**Part I — Single-cell QC**

1. [How this package thinks about QC](#1-how-this-package-thinks-about-qc)
2. [Look before you filter](#2-look-before-you-filter)
   - 2.1 [`PlotQCMetrics()` — the 30-second look](#21-plotqcmetrics--the-30-second-look)
   - 2.2 [`GenerateQCReport()` — the real report](#22-generateqcreport--the-real-report)
3. [Choosing cutoffs](#3-choosing-cutoffs)
4. [`ApplyQCFilters()` — applying them](#4-applyqcfilters--applying-them)
5. [`QCComparePlots()` — did the filter do what you meant?](#5-qccompareplots--did-the-filter-do-what-you-meant)
6. [Doublets — `calldoublet()`](#6-doublets--calldoublet)
7. [Batch effects — `BatchEffectQC()`](#7-batch-effects--batcheffectqc)
8. [Cell cycle — `assign_cell_cycle_phase()`](#8-cell-cycle--assign_cell_cycle_phase)
9. [Structural sanity checks](#9-structural-sanity-checks)

**Part II — Spatial QC**

10. [What changes when the data has coordinates](#10-what-changes-when-the-data-has-coordinates)
11. [Inspecting a spatial object](#11-inspecting-a-spatial-object)
12. [Metric-based QC still applies](#12-metric-based-qc-still-applies)
13. [Spatial artifacts](#13-spatial-artifacts)
    - 13.1 [Visium edge spots — `EdgeDetectionVisium()`](#131-visium-edge-spots--edgedetectionvisium)
    - 13.2 [Imaging-platform FOV edges — `detect_fov_edges()`](#132-imaging-platform-fov-edges--detect_fov_edges)
    - 13.3 [Tissue holes — `detect_tissue_holes()`](#133-tissue-holes--detect_tissue_holes)
14. [Is the annotation spatially coherent? — `SpatialConcordance()`](#14-is-the-annotation-spatially-coherent--spatialconcordance)
15. [Subsetting without corrupting images](#15-subsetting-without-corrupting-images)
16. [A spatial QC recipe](#16-a-spatial-qc-recipe)

---

# Part I — Single-cell QC

## 1. How this package thinks about QC

QC here is four separable questions, and it helps to keep them apart because
they fail differently:

| Question | Tool |
|---|---|
| Are these barcodes cells? | `GenerateQCReport()` → `ApplyQCFilters()` |
| Are they *single* cells? | `calldoublet()` |
| Is variation between samples technical? | `BatchEffectQC()` |
| Is variation within a sample a confounder? | `assign_cell_cycle_phase()` |

The metric-based path is deliberately **two-step**: a report that *suggests*
cutoffs and writes them to a CSV, then a separate call that *applies* them.
The point is that the CSV is editable. You are expected to look at the report,
disagree with some of its suggestions, change those numbers, and keep the file
as the record of what you actually did.

```{r, eval=FALSE}
library(SingleCellTools)
library(Seurat)

# objs is a named list of Seurat objects, one per sample
objs <- CreateRNAObjects(data_dirs, run_doublet_finder = TRUE)
```

## 2. Look before you filter

### 2.1 `PlotQCMetrics()` — the 30-second look

Violin panels of whatever QC columns exist, split by a grouping variable.
Use it to decide whether anything is obviously wrong before investing in the
full report.

```{r, eval=FALSE}
PlotQCMetrics(obj, group.by = "orig.ident")

# Restrict to specific metrics, control the grid
PlotQCMetrics(obj,
              group.by = "sample",
              qc_cols  = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
              ncol     = 3)
```

### 2.2 `GenerateQCReport()` — the real report

Renders a standalone HTML report and, crucially, writes a **sidecar CSV** of
suggested per-sample, per-metric cutoffs next to it
(`<report>_cutoffs.csv`). That CSV is the input to `ApplyQCFilters()`.

```{r, eval=FALSE}
GenerateQCReport(objs,
                 output_file = "qc/qc_report.html",
                 title       = "Experiment 1 QC",
                 sample_col  = "orig.ident")
```

Arguments worth knowing:

```{r, eval=FALSE}
GenerateQCReport(
  objs,
  output_file      = "qc/qc_report.html",
  metadata_cols    = c("nFeature_RNA", "nCount_RNA",
                       "percent.mt", "percent.rb", "percent.hb"),
  mad_multiplier   = 3,      # width of the suggested interval, in MADs
  log_skewed       = TRUE,   # log-scale metrics whose skew exceeds...
  log_threshold    = 10,     # ...this
  complexity_score = TRUE,   # genes-per-UMI, catches low-complexity barcodes
  doublet_col      = "doublet_finder"
)
```

`mad_multiplier` is the main dial. Cutoffs are median ± *k*·MAD per sample per
metric, so lowering it filters harder. MAD is used rather than SD because the
outliers you are trying to exclude would otherwise inflate the very statistic
defining the threshold.

## 3. Choosing cutoffs

Open the HTML, then open the CSV. It has one row per (sample, metric):

```
sample,metric,suggest_lo,suggest_hi,median,mad,pct_pass
Alt1,nFeature_RNA,214,3180,641,489,0.94
Alt1,percent.mt,0,8.7,1.2,2.1,0.96
...
```

Edit `suggest_lo` / `suggest_hi` freely — extra columns are ignored by the
filter and passed through. Some judgement:

- **`percent.mt` deserves a hard ceiling, not a MAD interval.** If a sample is
  uniformly high-mitochondrial, its median is high and MAD-based cutoffs will
  happily keep dying cells. Set an absolute value you're willing to defend.
- **A lower bound on `nCount_RNA` is doing cell-calling**, which the upstream
  pipeline already did. Be deliberate about overruling it.
- **Degenerate rows are skipped, not applied.** If a metric is constant across
  a sample its suggested quantiles coincide, and `suggest_lo == suggest_hi`
  would keep only cells matching that exact value. `ApplyQCFilters()` detects
  that (and missing or inverted bounds) and skips *that metric for that
  sample*, leaving the others in force. Watch for the message — it usually
  means the metric is uninformative for that sample.

## 4. `ApplyQCFilters()` — applying them

```{r, eval=FALSE}
objs_filt <- ApplyQCFilters(objs, cutoffs = "qc/qc_report_cutoffs.csv")
```

Point it at the `.html` and it finds the sidecar CSV itself. Pass a data frame
instead if you built cutoffs programmatically.

```{r, eval=FALSE}
res <- ApplyQCFilters(
  objs,
  cutoffs         = "qc/qc_report.html",   # sidecar resolved automatically
  metrics         = c("nFeature_RNA", "percent.mt"),  # subset of the table
  override        = list(percent.mt = c(0, 5)),       # beats the table
  filter_doublets = TRUE,
  return_report   = TRUE
)

res$obj        # filtered objects
res$report     # per-sample, per-metric retention
res$cutoffs    # what was actually applied
```

`return_report = TRUE` is worth making a habit. `res$report` gives
`n_before` / `n_after` / `pct_kept` per metric, which is how you notice that
one metric is doing all the filtering.

A single object with multiple samples in one column works too:

```{r, eval=FALSE}
ApplyQCFilters(merged, cutoffs = cut_df, sample_col = "sample")
```

## 5. `QCComparePlots()` — did the filter do what you meant?

Pre/post violins per sample with retention annotated.

```{r, eval=FALSE}
p <- QCComparePlots(objs, objs_filt)
p <- QCComparePlots(objs, objs_filt, metrics = c("nCount_RNA", "percent.mt"))
```

Read it for **shape**, not just counts. A filter that removed a tail is doing
its job; one that shifted the whole distribution removed a cell population.

Two practical notes:

- Objects in a list may carry per-sample metadata column *names* — most often
  DoubletFinder's `pANN_<pN>_<pK>_<nExp>`, whose values differ per sample.
  Columns not shared by every object are dropped with a warning, which is
  expected and harmless.
- `counts_size` shrinks the retention labels if they crowd each other on a
  panel with many samples.

## 6. Doublets — `calldoublet()`

A DoubletFinder wrapper. Adds `doublet_finder` (`"Singlet"`/`"Doublet"`) and
`doublet_pANN` (the raw score), and strips its own intermediates on the way
out.

```{r, eval=FALSE}
obj <- calldoublet(obj, samplenameIndex = 1)

# usually run through the loader instead
objs <- CreateRNAObjects(data_dirs,
                         run_doublet_finder = TRUE,
                         doublet_rate       = 0.075)
```

**`doublet_rate` is the argument that matters.** It is the assumed doublet
*formation* rate — the fraction of barcodes expected to be multiplets given
how many cells were loaded — and it sets how many doublets the classifier
looks for. It is deliberately **not** `pK`, which is the pANN neighborhood
size chosen from the data and carries no information about doublet abundance.

Scale it to your recovery rather than accepting the default. For 10X Chromium
the rate runs at roughly 0.8% per 1,000 cells recovered:

```{r, eval=FALSE}
# ~4,000 cells recovered -> ~3%
objs <- CreateRNAObjects(data_dirs, doublet_rate = 0.03)

# per-sample rates, if recovery varies a lot
objs <- Map(function(o, r) calldoublet(o, doublet_rate = r),
            objs, c(Alt1 = 0.035, Alt2 = 0.05, PBS1 = 0.04))
```

Speed knobs, since the pK sweep dominates the runtime:

```{r, eval=FALSE}
calldoublet(obj,
            pk_sweep_max_cells = 4000,  # estimate pK on a subsample
            sweep_cores        = 1)     # paramSweep's own parallelism
```

Raising `sweep_cores` above 1 while `CreateRNAObjects(workers = n)` is also
parallelizing oversubscribes the CPU; the loader errors up front if
`workers * doublet_sweep_cores` exceeds the core count.

Doublet filtering happens inside `ApplyQCFilters()` (`filter_doublets = TRUE`)
rather than at call time, so labels survive for review.

## 7. Batch effects — `BatchEffectQC()`

Quantifies how mixed your batches are in a reduction, so "did integration
work" stops being a matter of squinting at a UMAP.

```{r, eval=FALSE}
pre  <- BatchEffectQC(merged, reduction = "pca",     batch_col = "sample")
post <- BatchEffectQC(merged, reduction = "harmony", batch_col = "sample")

pre$summary
post$summary
```

Run it **before and after** integration. The absolute number means little; the
change means everything. With `celltype_col` set you also get `per_cluster`,
which is where over-integration shows up — a cluster that mixed *too* well may
have had a real biological difference erased.

```{r, eval=FALSE}
BatchEffectQC(merged, reduction = "harmony",
              batch_col = "sample", celltype_col = "celltype")
```

## 8. Cell cycle — `assign_cell_cycle_phase()`

UCell-based scoring, then a per-cell phase call. Works on human or mouse with
no species argument — gene symbols are matched to the object's features
case-insensitively, and the defaults come from
`Seurat::cc.genes.updated.2019`.

```{r, eval=FALSE}
obj <- assign_cell_cycle_phase(obj)
table(obj$Phase)
```

**Read `threshold_quantile` carefully.** A cell is G1 when *both* scores are at
or below their quantile cutoffs, so the default `0.5` calls roughly 60% of
cells "cycling" **on any dataset**, quiescent or not. It is a within-dataset
rank, not an absolute call.

Calibrate it against canonical markers rather than trusting the default:

```{r, eval=FALSE}
mk  <- intersect(c("Mki67", "Top2a", "Ccnb1", "Birc5"), rownames(obj))
det <- Matrix::colSums(LayerData(obj, layer = "counts")[mk, , drop = FALSE] > 0)

for (q in c(0.5, 0.75, 0.9, 0.95)) {
  p <- assign_cell_cycle_phase(obj, threshold_quantile = q)$Phase
  cat(sprintf("q=%.2f  cycling=%5.1f%%  marker+ among cycling=%5.1f%%\n",
              q, 100 * mean(p != "G1"), 100 * mean(det[p != "G1"] > 0)))
}
```

If precision barely improves as you tighten while recall collapses, the score
is only weakly discriminative at your depth — in which case **don't
discretize**. Regress on the continuous scores instead of the `Phase` factor:

```{r, eval=FALSE}
obj <- SCTransform(obj, vars.to.regress =
  c("signature_1_S.Score_UCell", "signature_1_G2M.Score_UCell"))
```

## 9. Structural sanity checks

Cheap, and they catch problems that look like biology later.

```{r, eval=FALSE}
# Duplicate gene symbols across a list of objects
check_duplicate_genes(objs)

# Gene ID type (symbol vs Ensembl vs Entrez) -- catches a mismatched reference
check_gene_ids_across_objects(objs)

# Per-cluster summary: sizes, QC metrics, sample composition
CellSuiteSummary(merged, cluster_col = "seurat_clusters", sample_col = "sample")
```

`CellSuiteSummary()` is the fastest way to spot a cluster that is one sample's
worth of low-quality cells wearing a cluster label.

---

# Part II — Spatial QC

## 10. What changes when the data has coordinates

Everything in Part I still applies — a low-count spot is still a low-count
spot. Spatial adds three things it cannot see:

1. **Position-dependent artifacts.** Tissue edges, FOV boundaries and holes
   produce systematically odd measurements that no per-cell metric flags.
2. **Images that must stay consistent with the cells.** Subsetting a spatial
   object with plain `subset()` can leave `@images` referencing cells that no
   longer exist.
3. **A spatial coherence question.** Is an annotation spatially structured, or
   scattered like noise?

## 11. Inspecting a spatial object

`SpatialObjectInfo()` returns one row per image — class, cell count, whether
an actual image is attached or only coordinates. Run it first; most spatial
confusion is really confusion about what `@images` contains.

```{r, eval=FALSE}
vis <- CreateVisiumObjects(data_dirs, file_type = "h5")
SpatialObjectInfo(vis)
```

Coordinates for every image in one data frame, optionally joined to metadata:

```{r, eval=FALSE}
coords <- get_all_coords(obj, meta_cols = c("seurat_clusters", "nCount_Spatial"))
head(coords)
```

That data frame is the escape hatch — anything the plotting helpers don't
cover, you can do in `ggplot2` directly from it.

## 12. Metric-based QC still applies

The same report/filter/compare loop works; only the metric names change.

```{r, eval=FALSE}
GenerateQCReport(vis,
                 output_file      = "qc/visium_qc.html",
                 metadata_cols    = c("nFeature_Spatial", "nCount_Spatial",
                                      "percent.mt"),
                 spatial_max_cols = 3)

vis_filt <- ApplyQCFilters(vis, cutoffs = "qc/visium_qc_cutoffs.csv")
QCComparePlots(vis, vis_filt,
               metrics = c("nCount_Spatial", "nFeature_Spatial"))
```

`QCComparePlots()` already log-scales `nCount_Spatial` / `nFeature_Spatial` by
default. `spatial_max_cols` controls the report's spatial panel grid.

One judgement call specific to spatial: a low-count spot at the tissue edge is
usually an artifact, while a low-count spot in the middle of dense tissue may
be a genuinely small or quiescent cell type. Filtering on counts alone treats
them identically, which is the argument for the next section.

## 13. Spatial artifacts

### 13.1 Visium edge spots — `EdgeDetectionVisium()`

Flags spots at the tissue border, where partial coverage depresses counts.
Note it returns a **data frame**, not a modified object — join it back
yourself.

```{r, eval=FALSE}
edges <- EdgeDetectionVisium(seurat.obj = vis[[1]],
                             search     = "radius",
                             radius_factor = 1.5)
head(edges)   # barcode, coordinates, Filter, Filter2, Filter3, Filter4
```

The four `Filter*` columns are increasingly aggressive definitions of "edge"
(`"Keep"` / `"Filter"`), so you can pick a stringency rather than accept one:

```{r, eval=FALSE}
vis[[1]]$edge <- edges$Filter2[match(colnames(vis[[1]]), edges$barcode)]
vis_core <- SubsetSpatial(vis[[1]], subset = edge == "Keep")
```

Neighbor search is either `"radius"` (spots within `radius_factor` × the
inferred spot spacing) or a fixed `neighbors` count. Radius is the safer
default on tissue with variable spot density.

### 13.2 Imaging-platform FOV edges — `detect_fov_edges()`

For Xenium/CosMx/MERFISH, where the artifact is the *imaging field* boundary
rather than the tissue border. Returns the object with an integer
`edge_layer` column: how many erosion passes in a cell sits.

```{r, eval=FALSE}
obj <- detect_fov_edges(obj, method = "bbox", n_iterations = 2)
table(obj$edge_layer)

# drop the outermost layer only
obj_core <- SubsetSpatial(obj, subset = edge_layer > 1)
```

`method = "bbox"` peels from the FOV bounding box and suits rectangular FOVs.
`"angular"` flags cells whose neighbors span less than `gap_threshold` radians
— cells on any boundary, including irregular tissue edges. Use `"angular"`
when FOVs are not rectangular or have been stitched.

### 13.3 Tissue holes — `detect_tissue_holes()`

Finds interior voids — tears, folds, vessel lumens — which read as
low-density regions rather than low-count cells.

```{r, eval=FALSE}
obj <- detect_tissue_holes(obj, min_hole_size = 4, sensitivity = 0.75)
table(obj$hole_layer)
```

Raise `sensitivity` to flag more aggressively. `min_hole_size` guards against
calling every small gap a hole. If a marker reliably marks the structure you
*don't* want called a hole (a lumen-lining gene, say), `exclude_gene` keeps it
out:

```{r, eval=FALSE}
detect_tissue_holes(obj, exclude_gene = "Pecam1",
                    exclude_gene_threshold = 1)
```

Both detectors add a label rather than removing cells, deliberately — inspect
the labels spatially before subsetting on them.

## 14. Is the annotation spatially coherent? — `SpatialConcordance()`

Tests whether cells sharing a label sit near each other more than chance, by
permutation. This is the closest thing to an objective check that a clustering
or deconvolution result is spatially real.

```{r, eval=FALSE}
SpatialConcordance(obj, group.by = "seurat_clusters", k = 6, n_perm = 100)
```

One row per image, with `observed` (mean same-label neighbor fraction),
`null_mean`, `z`, `p`, `padj`.

A high `z` says the labels are spatially structured. A `z` near zero says
they're scattered — which for a tissue-architecture cluster means the cluster
is probably technical. Run it on a shuffled label as a negative control:

```{r, eval=FALSE}
obj$shuffled <- sample(obj$seurat_clusters)
SpatialConcordance(obj, group.by = "shuffled")   # z should be ~0
```

`exclude` drops labels before scoring — useful for an "Unknown" or low-quality
class that would otherwise dominate.

## 15. Subsetting without corrupting images

**Use `SubsetSpatial()`, not `subset()`.** Plain subsetting can leave
`@images` inconsistent with the cells that remain; `SubsetSpatial()` subsets
every image alongside the expression data and drops images left with no cells,
with a message.

```{r, eval=FALSE}
obj_sub <- SubsetSpatial(obj, subset = celltype == "Hepatocyte")
obj_sub <- SubsetSpatial(obj, cells = keep_barcodes)
obj_sub <- SubsetSpatial(obj, idents = c("0", "3"))
obj_sub <- SubsetSpatial(obj, features = keep_genes)   # images untouched
```

Two related helpers:

```{r, eval=FALSE}
# Images named image, image.1, image.2... -> meaningful names
obj <- RenameSpatialImages(obj, group_col = "sample")
obj <- RenameSpatialImages(obj, new_names = c(image = "slice_A"))

# Shrink an object for sharing: drop images, or keep coordinates only
obj <- DropSpatialImage(obj, mode = "downgrade")   # coords stay, pixels go
obj <- DropSpatialImage(obj, mode = "remove")      # everything goes
```

`mode = "downgrade"` is usually what you want before `saveRDS()` — the hires
image is often most of the file size, and coordinates are what analysis needs.

## 16. A spatial QC recipe

```{r, eval=FALSE}
# 1. What am I holding?
vis <- CreateVisiumObjects(data_dirs, file_type = "h5")
SpatialObjectInfo(vis)
check_gene_ids_across_objects(vis)

# 2. Standard metric QC
GenerateQCReport(vis, output_file = "qc/visium_qc.html",
                 metadata_cols = c("nFeature_Spatial", "nCount_Spatial",
                                   "percent.mt"))
#    -> edit qc/visium_qc_cutoffs.csv here <-
vis_f <- ApplyQCFilters(vis, cutoffs = "qc/visium_qc_cutoffs.csv",
                        return_report = TRUE)
QCComparePlots(vis, vis_f$obj)

# 3. Spatial artifacts -- label, look, then decide
edges <- EdgeDetectionVisium(seurat.obj = vis_f$obj[[1]])
vis_f$obj[[1]]$edge <- edges$Filter2[match(colnames(vis_f$obj[[1]]),
                                           edges$barcode)]
SpatialDimPlotFixed(vis_f$obj[[1]], group.by = "edge")

# 4. Cluster, then ask whether the clusters are spatially real
merged <- MergeSeurat(vis_f$obj, spatial = "Visium")
SpatialConcordance(merged, group.by = "seurat_clusters")

# 5. Shrink before saving
saveRDS(DropSpatialImage(merged, mode = "downgrade"), "visium_qc.rds")
```

---

## See also

- The main [SingleCellTools vignette](SingleCellTools_vignette.html) for the
  full analysis workflow.
- `?GenerateQCReport`, `?ApplyQCFilters`, `?calldoublet` — every argument
  above is documented with its rationale.
